---
title: "Data Mining for Disaster Relief"
subtitle: "Classifying Blue Tarps in Post-Earthquake Haiti Satellite Imagery"
author: "Wyatt Scott"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# 1. Environment Setup

```{r load-utils}
# Source all helper functions (plotting, tables, data loading, etc.)
source(here::here("scripts", "utils.R"))

# Additional packages not loaded in utils.R
library(glmnet)
library(randomForest)
library(e1071)
library(kernlab)
library(MASS)
library(pROC)
library(plotly)
library(scatterplot3d)

# Start parallel backend
cl <- setup_parallel()
```

# 2. Data Ingestion

```{r load-data}
data    <- load_training_data()
holdout <- load_holdout_data()
```

```{r class-imbalance}
eda_table <- data.frame(
  `Tarp Status` = c("No", "Yes"),
  Training      = c("61,219", "2,022"),
  Holdout       = c("1,989,697", "14,480"),
  `Training %`  = c(96.8, 3.2),
  `Holdout %`   = c(99.27, 0.73),
  check.names   = FALSE
)

styled_kable(eda_table, caption = "<b>Table 1: Class Imbalance</b>")
```

# 3. Exploratory Data Analysis

```{r density-plots, fig.height=8}
# Training densities
p_red   <- plot_color_density(data, "Red",   "Red (Train)")
p_green <- plot_color_density(data, "Green", "Green (Train)")
p_blue  <- plot_color_density(data, "Blue",  "Blue (Train)")

# Holdout densities
p_b1 <- plot_color_density(holdout, "Red",   "Red (Holdout)")
p_b2 <- plot_color_density(holdout, "Green", "Green (Holdout)")
p_b3 <- plot_color_density(holdout, "Blue",  "Blue (Holdout)")

(p_b1 + p_red + p_b2 + p_green + p_b3 + p_blue) +
  plot_layout(guides = "collect", ncol = 2) +
  plot_annotation(
    title = "Figure 1: Color Distributions by Tarp Status"
  ) &
  theme(
    plot.title   = element_text(face = "bold", size = 12, hjust = 0.5),
    axis.ticks   = element_blank(),
    axis.title   = element_blank(),
    axis.text.y  = element_blank(),
    legend.position    = "bottom",
    legend.text        = element_text(face = "bold"),
    legend.title       = element_text(face = "bold"),
    legend.background  = element_rect(fill = "white", color = "grey35")
  )
```

```{r 3d-scatter}
# Subsample for visualization
set.seed(1)
data_sub    <- data    %>% sample_n(60000, replace = FALSE) %>% mutate(Source = "Training")
holdout_sub <- holdout %>% sample_n(60000, replace = FALSE) %>% mutate(Source = "Holdout")
combined    <- bind_rows(data_sub, holdout_sub)

plot_ly(
  combined,
  x = ~Green, y = ~Red, z = ~Blue,
  color  = ~interaction(Source, Tarp),
  colors = colorRamp(c("black", "skyblue", "purple", "#E57200")),
  type   = "scatter3d",
  mode   = "markers",
  marker = list(size = 1.6, opacity = 0.6, symbol = "circle-open")
) %>%
  layout(
    scene = list(
      xaxis = list(title = "Green"),
      yaxis = list(title = "Red"),
      zaxis = list(title = "Blue")
    ),
    title = "Figure 2: RGB Values by Dataset and Tarp Status"
  )
```

# 4. Model Training

All models use 10-fold stratified CV with AUC as the selection metric.

```{r cv-setup}
cv      <- make_cv_control(data)
ctrl    <- cv$ctrl
results <- list()   # Store model results for comparison
```

## 4.1 Logistic Regression

```{r logistic}
fit_logistic <- train_timed(Tarp ~ Red + Blue + Green, data, "glm", ctrl,
                            family = "binomial")
results$Logistic <- fit_logistic

thresh_log <- compute_thresholds(fit_logistic$model)
plot_threshold_grid(thresh_log, "Logistic")
plot_roc_curves(fit_logistic$model, data,
                title = "ROC Curves (Logistic)",
                xlim = c(0, 0.08), ylim = c(0.75, 1))
```

## 4.2 LDA

```{r lda}
fit_lda <- train_timed(Tarp ~ log(Red) + log(Blue) + log(Green), data, "lda",
                       ctrl, preProcess = c("center", "scale"))
results$LDA <- fit_lda

thresh_lda <- compute_thresholds(fit_lda$model)
plot_threshold_grid(thresh_lda, "LDA")
plot_roc_curves(fit_lda$model, data,
                title = "ROC Curves (LDA)",
                xlim = c(0, 0.15), ylim = c(0.85, 1))
```

## 4.3 QDA

```{r qda}
fit_qda <- train_timed(Tarp ~ Red + Blue + Green, data, "qda", ctrl)
results$QDA <- fit_qda

thresh_qda <- compute_thresholds(fit_qda$model)
plot_threshold_grid(thresh_qda, "QDA")
plot_roc_curves(fit_qda$model, data,
                title = "ROC Curves (QDA)",
                xlim = c(0, 0.15), ylim = c(0.85, 1))
```

## 4.4 KNN

```{r knn}
fit_knn <- train_timed(Tarp ~ Red + Blue + Green, data, "knn", ctrl,
                       tuneLength = 10)
results$KNN <- fit_knn

# Plot AUC vs. K
knn_k_df <- data.frame(
  K   = fit_knn$model$results$k,
  AUC = round(fit_knn$model$results$ROC, 5) * 100
)
ggplot(knn_k_df, aes(K, AUC)) +
  geom_line(color = "#232D4B", linewidth = 0.5) +
  geom_point(color = "#E57200", size = 1.5) +
  labs(title = "AUC by K", x = "K", y = "AUC (CV)") +
  scale_x_continuous(breaks = knn_k_df$K) +
  theme_linedraw() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"))

thresh_knn <- compute_thresholds(fit_knn$model)
plot_threshold_grid(thresh_knn, "KNN",
                    subtitle = paste0("k = ", fit_knn$model$bestTune$k))
plot_roc_curves(fit_knn$model, data,
                title = "ROC Curves (KNN)",
                note  = paste0("k = ", fit_knn$model$bestTune$k),
                xlim = c(0, 0.007), ylim = c(0, 1))
```

## 4.5 Penalized Logistic Regression (Elastic Net)

```{r plr}
fit_plr <- train_timed(Tarp ~ Red + Blue + Green, data, "glmnet", ctrl,
                       tuneLength = 20, allowParallel = TRUE)
results$PLR <- fit_plr

best_alpha  <- round(fit_plr$model$bestTune$alpha, 4)
best_lambda <- signif(fit_plr$model$bestTune$lambda, 2)
plr_note    <- paste0("\u03B1 = ", best_alpha, " | \u03BB = ", best_lambda)

thresh_plr <- compute_thresholds(fit_plr$model)
plot_threshold_grid(thresh_plr, "PLR", subtitle = plr_note)
plot_roc_curves(fit_plr$model, data,
                title = "ROC Curves (PLR)",
                note  = plr_note,
                xlim = c(0, 0.4), ylim = c(0.75, 1))
```

## 4.6 Random Forest

```{r rf}
mtry_grid <- expand.grid(mtry = c(1, 2))
fit_rf <- train_timed(Tarp ~ Red + Green + Blue, data, "rf", ctrl,
                      tuneGrid = mtry_grid, ntree = 500)
results$RF <- fit_rf

thresh_rf <- compute_thresholds(fit_rf$model)
rf_note <- paste0("ntree = 500 | mtry = ", fit_rf$model$bestTune$mtry)
plot_threshold_grid(thresh_rf, "Random Forest", subtitle = rf_note)
plot_roc_curves(fit_rf$model, data,
                title = "ROC Curves (RF)",
                note  = rf_note,
                xlim = c(0, 0.01), ylim = c(0, 1))
```

## 4.7 Support Vector Machine (RBF Kernel)

```{r svm}
svm_grid <- expand.grid(C = 0.25, sigma = seq(1, 3, 0.25))
fit_svm <- train_timed(Tarp ~ Red + Green + Blue, data, "svmRadial", ctrl,
                       tuneGrid = svm_grid)
results$SVM <- fit_svm

svm_note <- paste0(
  "C = ", fit_svm$model$bestTune$C,
  " | \u03B3 = ", round(fit_svm$model$bestTune$sigma, 4)
)

thresh_svm <- compute_thresholds(fit_svm$model)
plot_threshold_grid(thresh_svm, "SVM (RBF)", subtitle = svm_note)
plot_roc_curves(fit_svm$model, data,
                title = "ROC Curves (SVM)",
                note  = svm_note,
                xlim = c(0, 0.01), ylim = c(0, 1))
```

# 5. Model Comparison

```{r comparison-table}
comparison <- purrr::imap_dfr(results, function(res, name) {
  best_row <- res$model$results
  if (nrow(best_row) > 1) {
    # Pick the row matching bestTune
    best_idx <- as.numeric(rownames(res$model$bestTune))
    if (!is.na(best_idx) && best_idx <= nrow(best_row)) {
      best_row <- best_row[best_idx, ]
    } else {
      best_row <- best_row[which.max(best_row$ROC), ]
    }
  }
  data.frame(
    Model       = name,
    AUC         = round(best_row$ROC, 5) * 100,
    Sensitivity = round(best_row$Sens, 4) * 100,
    Specificity = round(best_row$Spec, 4) * 100,
    `Train Time (s)` = round(res$elapsed, 2),
    check.names = FALSE
  )
})

styled_kable(comparison, caption = "<b>Table: Cross-Validation Summary</b>")
```

# 6. Holdout Evaluation

```{r holdout-eval}
# Deploy each model on the holdout set, measure prediction time
holdout_results <- purrr::imap_dfr(results, function(res, name) {
  start <- proc.time()
  preds <- predict(res$model, newdata = holdout)
  elapsed <- (proc.time() - start)["elapsed"]

  cm <- confusionMatrix(preds, holdout$Tarp, positive = "Yes")

  data.frame(
    Model             = name,
    `Balanced Accuracy` = round(cm$byClass["Balanced Accuracy"], 4) * 100,
    Kappa             = round(cm$overall["Kappa"], 4) * 100,
    F1                = round(cm$byClass["F1"], 4) * 100,
    FPR               = round(1 - cm$byClass["Specificity"], 4) * 100,
    `Deploy Time (s)` = round(elapsed, 2),
    check.names       = FALSE
  )
})

styled_kable(holdout_results, caption = "<b>Table: Holdout Performance</b>")
```

```{r holdout-confusion-matrices, fig.height=4, fig.width=10}
cm_plots <- purrr::imap(results, function(res, name) {
  preds <- predict(res$model, newdata = holdout)
  cm    <- confusionMatrix(preds, holdout$Tarp, positive = "Yes")
  plot_confusion_matrix(cm, name)
})

do.call(gridExtra::grid.arrange, c(cm_plots, ncol = 4))
```

# 7. Conclusions

The **Penalized Logistic Regression** model is recommended for blue tarp detection.
On the holdout set it achieved the highest AUC, Kappa, and F1 score, the
second-lowest FPR, and the fastest deployment time. The traditional Logistic
and LDA models are close contenders — their simplicity and interpretability
make them strong alternatives when communicating results to non-technical
stakeholders.

Key limitations include the high multicollinearity among RGB predictors
(VIFs > 200), the severe class imbalance, and the use of a single geographic
region for training. Future work should explore color space transformations
(e.g., HSL or CIELUV), over-sampling techniques like SMOTE, and validation
on imagery from other disaster sites.

```{r cleanup}
stopCluster(cl)
```

# Session Info

```{r session-info}
sessionInfo()
```
